---
title: "DATA608 - Homework One - Principles of Data Visualization and Introduction to ggplot2"
author: "Chris G Martin"
date: "February 5, 2017"
fontsize: 11pt
output:
  html_document:
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: no
    theme: united
    toc: yes
  pdf_document:
    fig_caption: yes
    highlight: pygments
    latex_engine: xelatex
    number_sections: no
    toc: yes
---

#Objective

Using a dataset about the 5,000 fastest growing companies in the US, as compiled by Inc. magazine (limited to ggplot2):

```{r echo=FALSE, message=FALSE, warning=FALSE}
#load packages
library(dplyr)
library(ggplot2)
#install.packages('choroplethr')
library(choroplethr)
#install.packages('choroplethrMaps')
library(choroplethrMaps)

```

```{r echo=FALSE}
inc500 <- read.csv("https://raw.githubusercontent.com/chrisgmartin/DATA608/master/lecture1/inc5000_data.csv", stringsAsFactors=FALSE)
```

##Question One

Create a graph the at shows the distribution of companies in the dataset by State (ie how many are in each state). There are a lot of States, so consider which axis you should use assuming I am using a 'portrait' oriented screen (ie taller than wide).

###Variety of Charts/Graphics

```{r echo=FALSE}
#Count number of companies per state
statecount <- inc500 %>% 
  group_by(., State) %>% 
  dplyr::summarise(count = n())
```


####Map of the United States

```{r echo=FALSE, warning=FALSE}
#US Map
stateConversion <- function(x, faclevs = 'selected') {
 
    st.codes <- data.frame(state = as.factor(c("AK", "AL", "AR", "AZ", "CA", "CO", "CT", "DC", "DE", "FL", "GA", "HI",
                                            "IA", "ID", "IL", "IN", "KS", "KY", "LA", "MA", "MD", "ME", "MI", "MN",
                                            "MO", "MS",  "MT", "NC", "ND", "NE", "NH", "NJ", "NM", "NV", "NY", "OH",
                                            "OK", "OR", "PA", "PR", "RI", "SC", "SD", "TN", "TX", "UT", "VA", "VT",
                                            "WA", "WI", "WV", "WY")),
                            full = as.factor(c("Alaska","Alabama" ,  "Arkansas", "Arizona","California" , "Colorado" ,
                                            "Connecticut", "District of Columbia","Delaware" ,  "Florida" , "Georgia" ,
                                            "Hawaii","Iowa" ,"Idaho" , "Illinois" , "Indiana" ,  "Kansas" ,
                                            "Kentucky" , "Louisiana" , "Massachusetts", "Maryland" ,"Maine" ,
                                            "Michigan" , "Minnesota" , "Missouri" ,"Mississippi" ,  "Montana" ,
                                            "North Carolina","North Dakota", "Nebraska" , "New Hampshire" , "New Jersey" ,  "New Mexico" ,
                                            "Nevada" ,"New York" , "Ohio" , "Oklahoma" ,
                                            "Oregon" , "Pennsylvania" , "Puerto Rico", "Rhode Island" , "South Carolina", "South Dakota" ,
                                            "Tennessee" , "Texas" , "Utah" ,  "Virginia","Vermont" ,
                                            "Washington" , "Wisconsin", "West Virginia" , "Wyoming"))
    )
 
    if (nchar(x[1]) == 2) { st.x <- data.frame(state = x); refac.x <- st.codes$full[match(tolower(st.x$state), tolower(st.codes$state))] }
    else { st.x <- data.frame(full = x); refac.x <- st.codes$state[match(tolower(st.x$full), tolower(st.codes$full))] }
 
    if(faclevs == 'all') {return(refac.x)}
    else {return(factor(refac.x))}
 
}

statecount2 <- statecount
statecount2$State <- tolower(stateConversion(statecount$State, faclevs = 'selected'))
colnames(statecount2) <- c('region','value')
state_choropleth(statecount2, title = "Fastest Growing Companies (excl. Puerto Rico)", legend = "Num of Companies", num_colors = 8)
```

####Scatterplots

Did not think a scatterplot was appropriate.

####Bar Charts

```{r echo=FALSE, warning=FALSE}
statecount2 <- statecount
statecount2$xaxis <- 1

ggplot(statecount2, aes(x=State, y=count, fill=State)) + geom_bar(stat="identity", colour="gray32", width=0.4, show_guide=FALSE) + ggtitle("Count of Fastest Growing Companies")
```

```{r echo=FALSE}
statecount2 <- statecount

y.breaks <- cumsum(statecount2$count) - statecount2$count/2

ggplot(statecount2, aes(x=1, y=count, fill=State)) + geom_bar(stat="identity") + ggtitle("Continuous Count of Fastest Growing Companies") + geom_bar(stat="identity", colour='black') + guides(fill=guide_legend(override.aes=list(colour=NA))) + theme(axis.text.x=element_text(color='black')) + scale_y_continuous(breaks=y.breaks, labels=statecount2$State)
```

####Circular Charts

Note, I do agree with the sentiment that pie charts are not always the best visual aides. I try not to use them but I do like to use them in the process of finding the best visual aide. This section is not limited to pie charts, but general circular charts (such as radar).

```{r echo=FALSE}
statecount2 <- statecount

ggplot(statecount2, aes(x=1, y=count, fill=State)) + geom_bar(stat="identity") + ggtitle("Count of Fastest Growing Companies in the United States") + coord_polar(theta = 'y')
```


```{r echo=FALSE}
statecount2 <- statecount

ggplot(statecount2, aes(x=count, fill=State)) + coord_polar() + geom_bar(width=1) + ggtitle("Count of Fastest Growing Companies")
```


###Recommended/Favorite Graphic

My personal favorite graphic was the first bar chart, but it does not fit the display very well. So perhaps teh best chart is the continuous bar chart. The chart I'd like to use, I could not figure out how to create: a single line on the y-axis stretching from 0 to the maximum state count, with a label at the count per each state.



###Question Two

Let's dig in on the State with the 3rd most companies in the data set. Imagine you work for the state and are interested in how many people are employed by companies in different industries employ. Create a plot of average employment by industry for companies in this state (only use cases with full data (user R's complete.cases() function). Your graph should show how variable the ranges are, and exclude outliers.

```{r echo=FALSE, warning=FALSE}
#filter for third highest state count
statecount3 <- statecount
statecount3 <- statecount3[order(-statecount3$count),]
statecount3 <- statecount3$State[3]
indcount <- inc500[complete.cases(inc500),]
indcount <- filter(indcount, State == statecount3)

ggplot(indcount, aes(x=Industry, y=Employees)) + geom_boxplot(outlier.shape = NA) + scale_y_continuous(limits=quantile(indcount$Employees, c(0.1, 0.9))) + theme(axis.text.x = element_text(angle=90, hjust=1)) + ggtitle("Employees per Industry")
```


###Question Three

Now imagine you work for an investor and want to see which industries generate the most revenue per employee. Create a chart makes this information clear.

```{r echo=FALSE, warning=FALSE}
#all of the United States
revemp <- inc500
revemp <- revemp[complete.cases(revemp),]
revemp$RevperEmp <- revemp$Revenue/revemp$Employees
ggplot(revemp, aes(x=Industry, y=RevperEmp)) + geom_boxplot(outlier.shape = NA) + scale_y_continuous(limits=quantile(revemp$RevperEmp, c(0.1, 0.9))) + theme(axis.text.x = element_text(angle=90, hjust=1)) + ggtitle("Revenue per Employees per Industry (all of the United States)")


#for third highest state count
revemp <- inc500[complete.cases(inc500),]
revemp <- filter(revemp, State == statecount3)
revemp$RevperEmp <- revemp$Revenue/revemp$Employees
ggplot(revemp, aes(x=Industry, y=RevperEmp)) + geom_boxplot(outlier.shape = NA) + scale_y_continuous(limits=quantile(revemp$RevperEmp, c(0.1, 0.9))) + theme(axis.text.x = element_text(angle=90, hjust=1)) + ggtitle("Revenue per Employees per Industry (Third Higest State with FGC)")
```





##All Together (Code Displayed)

```{r eval=FALSE}
#load data
inc500 <- read.csv("https://raw.githubusercontent.com/chrisgmartin/DATA608/master/lecture1/inc5000_data.csv", stringsAsFactors=FALSE)

#QUESTION ONE
#Count number of companies per state
statecount <- inc500 %>% 
  group_by(., State) %>% 
  dplyr::summarise(count = n())

#using a second df for data quality
statecount2 <- statecount

#continuous count bar chart
#for US state counts
y.breaks <- cumsum(statecount2$count) - statecount2$count/2

ggplot(statecount2, aes(x=1, y=count, fill=State)) + geom_bar(stat="identity") + ggtitle("Continuous Count of Fastest Growing Companies") + geom_bar(stat="identity", colour='black') + guides(fill=guide_legend(override.aes=list(colour=NA))) + theme(axis.text.x=element_text(color='black')) + scale_y_continuous(breaks=y.breaks, labels=statecount2$State)

#QUESTION TWO
#filter for third highest state count and remove incomplete cases
statecount3 <- statecount
statecount3 <- statecount3[order(-statecount3$count),]
statecount3 <- statecount3$State[3]
indcount <- inc500[complete.cases(inc500),]
indcount <- filter(indcount, State == statecount3)

#plot
ggplot(indcount, aes(x=Industry, y=Employees)) + geom_boxplot(outlier.shape = NA) + scale_y_continuous(limits=quantile(indcount$Employees, c(0.1, 0.9))) + theme(axis.text.x = element_text(angle=90, hjust=1))

#QUESTION THREE
#all of the United States
revemp <- inc500
revemp <- revemp[complete.cases(revemp),]
revemp$RevperEmp <- revemp$Revenue/revemp$Employees
ggplot(revemp, aes(x=Industry, y=RevperEmp)) + geom_boxplot(outlier.shape = NA) + scale_y_continuous(limits=quantile(revemp$RevperEmp, c(0.1, 0.9))) + theme(axis.text.x = element_text(angle=90, hjust=1)) + ggtitle("Revenue per Employees per Industry (all of the United States)")


#for third highest state count
revemp <- inc500[complete.cases(inc500),]
revemp <- filter(revemp, State == statecount3)
revemp$RevperEmp <- revemp$Revenue/revemp$Employees
ggplot(revemp, aes(x=Industry, y=RevperEmp)) + geom_boxplot(outlier.shape = NA) + scale_y_continuous(limits=quantile(revemp$RevperEmp, c(0.1, 0.9))) + theme(axis.text.x = element_text(angle=90, hjust=1)) + ggtitle("Revenue per Employees per Industry (Third Higest State with FGC)")
```